Modelling last glacial cycle ice dynamics in the Alps

### 2018-07-23 articles/cycle/response-referee3.tex

    We reworked the first sentence. It now reads:

    \msquote{%
        During MIS 2, all six simulations yield comparative cumulative
        glaciated areas modelled maximum extents (Fig.~3a).}

    \referee{\textbf{p.~11, l.~10:}
        Is it possible some ice is missing from the geological reconstruction
        in some instances? I guess some outlets are well constrained, whilst
        others areas could be ``filled in'' by this modelling experiment.}

    Yes, parts of the reconstructed LGM ice limits are uncertain due to the
    lack of moraines or dated evidence. However, the discrepancies in ice
    extent discussed here are beyond uncertainties.

@@

% ----------------------------------------------------------------------

\sechead{Section 4}

    \referee{\textbf{p.~14, l.~15--31:}
        I find this description of sites and timing of glacier extent compared
        to dates difficult to follow. I suggest a new figure to convey this
        important comparison: Have the reconstructed and modelled ice extents
        at key times for each of the mentioned glaciers on several smaller
        maps, including geochronological constraints.}

    This would be very useful but to our knowledge, no geological data have
    been made available that would allow such a comparison. The review by
    \citep[Fig.~5]{Wirsig.etal.2016} which we refer to is the most up-to date
    compilation of LGM ages in and around the Alps, and deglacial outlines like
    there exists for the Eurasian and Laurentide ice sheets are not available
    for the Alps.

@@

    We added a reference to the study by \citet{Kuchar.etal.2012} although this
    approach is not doable for the Alps.

    \msquote{%
        Unfortunately, validation through the bedrock uplift rates
        \citep[cf.][]{Kuchar.etal.2012} is not doable in the Alps due its lower
        values, active tectonics and uncertainties on geological properties
        \citep[cf.][]{Mey.etal.2016}.}

    \referee{%
        A philosophical but important point is that your discussion throughout
        is written from the standpoint of the geochronological/geomorphological
        data and reconstructions as being ``truth''. It should consider
        somewhere that perhaps data is missing as it is hard-won in the places
        it exists and interpretations may be slightly wrong. The model is also
        not the ``truth'' and there is probably a blurred line in-between upon
        which we can proceed.}

    Thank you. Indeed the quality and availability of geomorphological and
    geochronological data is very variable across the Alps, and long-standing
    interpretations are sometimes proven wrong
    \citep[e.g.,][]{Monegato.etal.2007}. We appreciate your opinion, but the
    parts of the Alpine community sees our challenging of the trimline
    assumption as quite provocative given that it has been around for at least
    a century \citep{Penck.Bruckner.1909}.

### 2018-07-23 articles/cycle/response-referee1.tex

    Thank you very much for your detailed review of our manuscript.

    \referee{%
        This paper is a landmark advance in modelling European Alpine ice
        cover, applying a high-resolution (1\,km) ice model to the entire Alps
        through the last glacial cycle, for the first time to my knowledge.
        Results are compared with diverse geological data, and several
        important findings are presented, including time-transgressive ice
        marginal extents at LGM. The climate forcing is simple, applying
        uniform perturbations to modern observed datasets, which leads to some
        uncertainty in the results, but does not detract from them too much
        given the advances made in the ice modelling alone.}

    \referee{%
        The introduction gives an elegant summary of Alpine glacial science
        since the 1700's, including many historical references. The paper is
        well organized, with well-chosen sensitivities described first that
        calibrate the climate forcing, followed by detailed analysis of one
        best-fit high-resolution (1\,km) simulation through the last 120\,kyrs.
        Detailed comparisons to a variety of geological data are made,
        constituting a thorough assessment of model performance. An impressive
        animation of the whole cycle is included as supplementary material.}

    Thank you very much for these elogious and supportive comments!

@@

% ----------------------------------------------------------------------

\sechead{Specific comments}

    \referee{\textbf{p.~4, l.~9--10:}
        Can the physical basis of englacial water fraction and sensitivity of
        results be summarized briefly? This is not a usual component in
        ice-sheet models. Is the cap value ("capped at 0.01") well constrained,
        and does it have a significant effect on results?}

    Thank you for bringing this up.
    Laboratory experiments have demonstrated that the rheology of temperate,
    polycrystalline ice depends on its content in liquid water
    \citep[p.~65--66]{Cuffey.Paterson.2010}. Unfortunately, the only
    measurements available to date \citep{Duval.1977}, used to quantify the
    effect of liquid water on ice softness, the creep parameter $A$ in Glen's
    flow law \citep{Lliboutry.Duval.1985}, only extend to fractions of liquid
    water content between 0 and 0.8\%. They show a three-fold increase of ice
    softness over this range \citep[Fig.~1]{Duval.1977}.

    Ice sheet models have previously ignored this effect, but it has now
    become a typical component of polythermal models such as SICOPOLIS
    \citep{Greve.1997}, COMICE \citep{Ruckamp.etal.2010}, PISM
    \citep{Aschwanden.etal.2012}, ISSM \citep{Seroussi.etal.2013}, and
    TIM-FD$\unit{^{3}}$ \citep{Kleiner.Humbert.2014}.

    However water fractions between 1 and 5\,\% have repeatedly been observed
    in temperate glaciers \citep{Murray.etal.2000, Murray.etal.2007,
    Bradford.Harper.2005, Bradford.etal.2009}, and also occur in model results
    \citep[e.g.,][]{Blatter.Greve.2015}, but it is not known whether ice
    viscosity continues to decrease substancially for values above 0.8\,\%.
    Previous modelling studies have commonly assumed constant ice viscosity
    above 1\,\%. This arbitrary threshold is not constrained at all, and the
    urgent need for new ice deformation experiments has already been pointed
    out \citep{Kleiner.etal.2015}.

    The sensitivity of our results to the 1\,\% threshold was not tested.
    However, in our model results, liquid water fractions above 1\,\% typically
    only occur within the basal temperate layer of the fastest-moving glaciers
    where ice movement is largely dominated by basal sliding. We suspect that
    increased ice deformation in these regions is negligible in comparaison to
    uncertainties related to basal sliding and, more importantly, to climate 
    forcing.

    Thus, we prefer to avoid including the above discussion in the manuscript,
    but have reworked the sentence on water content:

    \msquote{%
        [Ice softness] increases with liquid water fractions up to 0.01
        \citep[p.~65--66]{Duval.1977, Lliboutry.Duval.1985,
        Cuffey.Paterson.2010}, an arbitrary threshold above which new
        ice deformation measurements are critically needed
        \citep{Kleiner.etal.2015}.}

@@

    The uncertainty to unknown rheology of water-rich temperate ice was also
    mentioned in the conclusions:

    \msquote{%
        In the absence of ice deformation measurements, a constant rheology was
        used for temperate ice containing more than 1\,\% of liquid water.}

@@

    \referee{\textbf{p.~4, l.~19--20:}
        The sub-glacial hydrologic component should be described more (even if
        exactly as in Bueler and van Pelt, 2015). Is basal water transported
        horizontally down the hydropotential gradient? This is usually a highly
        uncertain component of ice-sheet models, but can have a large effect on
        results through its influence on basal sliding, and basal frozen vs.
        thawed areas, which is relevant to section 4.4 regarding trimlines.}

    We refer to \citet{Bueler.Pelt.2015} as their paper contain the most
    up-to-date description of PISM till effective pressure physics used in our
    simulations \citep[Eqs.~18, 23, and 24]{Bueler.Pelt.2015}. However,
    subglacial water is not routed down the hydropotential gradient. We have
    clarified this:

@@

    \msquote{%
        Effective pressure is related to the ice overburden stress and the
        modelled amount of subglacial water, using a~formula derived from
        laboratory experiments with till extracted from the base of Ice Stream
        B in West Antarctica \citep[Table~1;][Eqs~23 and
        24]{Tulaczyk.etal.2000, Bueler.Pelt.2015}. Basal meltwater is
        accumulated locally without transportation. When the till becomes
        saturated, additional meltwater assumed to drain off instantaneously
        outside the glacier margins, i.e. it is removed from the system in an
        accountable way.}

    \referee{\textbf{Somewhat related:}
        Little information is given on the choices of basal sliding parameter
        values in Table 1. This could be discussed briefly. Presumably no
        inversion or optimization was performed for these values beforehand,
        and they do vary spatially. Are they appropriate for Alpine bedrock
        overall?}

    Although we refrain from repeating parameter values given in Table~1 in the
    main text, the following text was added in the methods:

@@

    Inversion of specific basal sliding parameters for past Alpine glaciers
    is difficult because the altitude and age of maximum ice surface elevation
    is discussed (cf. introduction and discussion on trimlines), and also
    depends on the even more uncertain regional climate history.  Therefore,
    basal sliding was also mentioned as one of the major sources of
    uncertainty in the conclusions.

    \msquote{%
        The till deformation model used here does not hold for sliding over
        bedrock surfaces. On the other hand, the constant friction angle used
        is representative of wet till but weaker basal conditions may have
        applied over saturated lake sediments where they occured.}

    \referee{\textbf{p.~7, l.~7:}
        Are there any data to support this atmospheric lapse rate value
        (6\,K\,km$^{-1}$), and do other values have the potential to
        significantly affect ice temperatures? In particular, could they change
        the basal areas of frozen/unfrozen ice and so the comparisons with
        trimlines in section 4.4?}

    In the European Alps, monthly temperature lapse rates vary between
    approximately 4\,K\,km$^{-1}$ in winter and 7\,K\,km$^{-1}$ in summer, and
    annual temperature lapse rates vary spatially between 5.4 and
    5.8\,K\,km$^{-1}$ \citep{Rolland.2003}. A reference to the study by
    \citet{Rolland.2003} was added in Table~1, and as we have now explicited,
    our constant value of 6\,K\,km$^{-1}$ is thus:

@@

    \msquote{%
        slightly above average but more representative of summer months when
        surface melt occurs \citep[Fig.~3]{Rolland.2003}.}

    Although no sensitivity tests were conducted, we argue here that the effect
    of atmospheric temperature lapse rate variations on ice temperature is
    negligible. First, seasonal variations do not penetrate ice or bedrock
    below a few metres. Second, even above 2\,km of ice, spatial variations of
    0.4\,K\,km$^{-1}$ would translate into surface temperature variations of
    only 0.8\,K. But during the Last Glacial Maximum, ice surface temperatures
    in the trimline region are typically between 15 and 20\,K below freezing.
    Thus the effect on the temperature gradient would be small.  Actually, the
    effect would even nearly disappear near the glacier base, where the
    temperature gradient is much steeper and primarily controlled by geothermal
    heat flux and shear heating.

@@

% ----------------------------------------------------------------------

\sechead{Climate forcing}

    \referee{%
        The method of spatially uniform shifts to modern climate forcing is
        common in paleo-modeling of large ice sheets, and in my opinion is
        acceptable as a starting point in this work, with coupling to regional
        climate models (RCMs) left to follow-on work. There are good
        discussions on possible shortcomings of this method, for instance as a
        cause of anomalous east-west marginal ice extents at LGM (pg. 11, line
        6-8). However, I suggest changing the sentence on pg. 9, line 5, which
        mentions some RCMs applied to LGM Europe, but also says "...over the
        Alps during the last glacial cycle, of which little is known apart from
        the LGM". There are several other RCM modeling studies over Europe
        during the last 120\,kyrs, e.g., for MIS Stage 3, Kjellstrom et al.,
        Boreas, 2010; Barron and Pollard, Quat. Res., 2002; Alfano et al.,
        Quat. Res., 2002; and for 6\,ka, Strandberg et al., Clim. Past, 2014.
        Perhaps there is little useful material there for the Alps, but such
        papers exist.}

    Thank you.
    We did not know about references on MIS~3 and have developped our sentence
    to better reflect the current state of knowledge on palaeo-precipitation:

    \msquote{%
        Palaeoclimate proxies indicate slightly reduced LGM precipitation in
        western Europe with anomalies diminishing eastwards
        \citep{Wu.etal.2007}. Regional circulation models indicate generally
        dryer conditions during MIS~3 \citep{Barron.Pollard.2002,
        Kjellstrom.etal.2010} but more precipitation south of the Alps during
        MIS~2 \citep{Strandberg.etal.2011, Ludwig.etal.2016}.}

@@

    \referee{%
        The past climate variations are prescribed following 3 quite distal
        core records, and the most distal (EPICA) is chosen as yielding the
        best fit to Alpine glacial evidence. The basis for preferring EPICA
        seems reasonable (matching some higher-frequency amplitudes of ice
        variability, section 3.3). However, this agreement is in a sense
        coincidental, in that there is no direct meteorological link between
        Antarctic and Alpine regional climate variations. Are there any
        proximal proxy records of Alpine climate at all, perhaps lacustrian
        varves, that could be used to assess the EPICA-based shifts in air
        temperatures and precipitation, even over limited periods of the last
        120\,kyrs?}

    A short review of available proximal proxy records was added here:

    \msquote{%
        Only few regional proxy records exist that extend over periods when the
        Alps were glaciated \citep{Heiri.etal.2014}. These include lake
        sediment records in north and west of the Alps
        \citep[e.g.,]{Beaulieu.Reille.1992, Wohlfarth.etal.2008,
        Duprat-Oualid.etal.2017}, and cave speleothems in the Eastern
        \citep[e.g.,][]{Spotl.Mangini.2002, Boch.etal.2011} and Western
        \citep{Luetscher.etal.2015} Alps. Due to the scarcity of vegetation
        north of the Alps during glacial periods, varying sources for moisture
        advection, and the limited duration of the records, quantitative
        palaeoclimatic interpretation will require combining multiple proxies
        in space and time, and comparing them against regional circulation
        model output \citep{Heiri.etal.2014}.}

@@

    \referee{%
        The point is well taken that trimlines do not necessarily indicate past
        ice surface elevations, but the upper limit of temperate ice with cold
        ice above (pg. 18, line 5 to pg. 19, line 4). It is an important point,
        because model LGM ice surfaces are far above most trimline elevations
        (as noted and in Fig.~6a,b). The pertinent results are shown in
        Fig.~6c, and I agree there is good support for the cold vs. temperate
        (basal) ice hypothesis.}

    Thank you for emphasizing this point.

    \referee{%
        It might help general readers to spell out the interpretation even more
        in the text. That is, as I understand it, the observed trimlines should
        coincide with the boundaries between model areas of frozen vs.
        temperate beds, so the dots in Fig.~6c should all lie on the borders
        between the hatched and white areas. The sentence on p.~19, l.~20-21,
        is confusing in this regard. Incidentally, it would also help to add
        the word "basal" to the last sentence of the Fig.~6c caption:
        "...experienced temperate basal ice for ...".}

    We realise that Fig.~6 is confusing. We had chosen the 1\,ka limit because temperate
    basal ice tends to occur above the trimlines during short periods of warmer
    climate. To avoid confusion, Fig.~6c has been simplified to show the basal
    thermal boundary at the age of maximum ice thickness. The last sentence in
    the caption now reads:

@@

    \msquote{%
        In the upper Rhone Valley, observed trimlines are often located near
        the LGM cold-temperate basal thermal transition, or in cold-based areas
        (Fig.~6c).}

    \referee{%
        One reason for the remaining discrepancies in Fig.~6c could be temporal
        variations in the model boundaries, that are aggregated in time by the
        "< 1\,kyr" criterion for the hatching and the grouping of all trimline
        data. To go into this in more detail, in principle Fig.~6c could be
        expanded to show the model basal frozen-temperate boundaries at
        particular times (21.5, 22.5, etc, ka), with only the dots for each
        time period superimposed. But that may not be worth it unless there are
        large temporal variations in the model boundaries.}

    Fig.~6c has been simplified to show the basal thermal boundary at the age
    of maximum ice thickness. Although this results in a time-transgressive
    picture, it means that the frozen areas indicated on the new figure are
    contemporaneous with the plotted ages and surface topography contours, and
    data shown on panels~a and b.

    Although the model output allows to study the migration of the basal
    thermal boundary over time into more detail, the basal velocity, which also
    controls erosion, should probably be considered as well. We leave this for
    future studies. The following sentence was added in the main text:

    \msquote{%
        The remaining discrepancies may relate to temporal migrations of the
        basal thermal boundary, an absence of sliding in warm-based areas [...]}

@@

    \referee{%
        A slight concern is that the majority of the trimline data seems to be
        orange dots i.e., older that 27\,ka in the timescale of Fig.~6c. The
        period for the model hatching extends back only to 29\,ka. Hopefully,
        most of the orange-dotted data are within that period and are not older
        than 29\,ka(?).}

    In fact, much of the mountainous areas reach maximum ice thickness during
    early MIS~2 when ice is colder and stiffer. This is also visible on
    Fig.~5. Some areas even reach maximum ice thickness during MIS~4 when the
    bedrock is slightly less depressed. There is certainly very much scope to
    discuss the age of the trimlines, which may differ significantly from that of
    the maximum ice extent on the lowland, but we are not aware of any data
    that could be used for validation here. However, frozen areas plotted on
    Fig.~6c are now consistent with the modelled ages used in panels a and c.

    \referee{%
        The text could briefly mention (and hopefully rule out) the issue of
        very fine-scale topographic features on which the trimlines are
        located, not resolved by the 1-km model topography. If data sites are
        on small-scale highs or lows significantly different from their
        \~km-scale surroundings, that could contribute to the discrepancies in
        Fig.~6c.}

    This is a valid point which we can unfortunately not rule out. Despite
    PISM's enthalpy scheme, which ensures a seamless transition between cold
    and temperate ice physics, resolving the basal thermal boundary on the
    valley sides is delicate, because it implies resolving a steep (vertical)
    enthalpy gradient over a steep bedrock slope. This requires both high
    vertical and high horizontal grid resolutions. This limitation was mentioned
    in the section on trimlines:

@@

    \referee{\textbf{p.~3, l.~2:}
        Perhaps change ``lead'' to ``led'', ``to which'' to ``to what''.}

        Done.

    \referee{\textbf{Fig.~1 caption, l.~4:}
        Perhaps change ``estimated'' to ``estimate'', or ``estimated of'' to
        ``estimated''.}

        Done. We thank you very much again for the time and effort you put into
        our manuscript.

### 2018-07-23 articles/cycle/response-referee2.tex

    \referee{%
        This manuscript describes a model study with the Parallel Ice Sheet
        Model applied to the last glacier cycle in the Alps. The climate
        forcing is derived from present day climate of WorldClim and the
        ERA-Interim reanalysis and time-dependent temperature offsets derived
        from the Greenland Ice core (GRIP), from Antarctic ice core (EPICA) and
        Marine sediment core from the Iberian margin (MD01-2444). The study is
        split in two, in the first part analysis of six model simulations (with
        and without precipitation scaling) made on a 2 km resolution grid is
        presented and concluded that out of the three climate forcing records
        used, the EPICA record gives the most realistic ice volume history
        during MIS 4 and 2. The second part analyses simulation made with the
        EPICA forcing record on a 1km grid. The authors draw conclusions about
        the ice cover, ice flow pattern, ice thickness, LGM ice extent, which
        in their model is a transient stage with varying timing across their
        model domain due to glacier dynamics.}

    Thank you very much for this fully accurate summary.

    \referee{%
        The manuscript is well organized
        and clearly written, the missing thing in this study is a discussion of
        and preferably a sensitivity study of the ice dynamic-model assumptions
        made. Is the sliding of the ice age ice sheet realistically modelled
        with pseudo-plastic assumption using the Shallow Shelf approximation?
        How sensitive are the results to the selected model parameters? It is
        briefly mentioned once on page 19, line 8, but a thorough analysis of
        the model sensitivity would strengthen the paper.}

    Thank you.
    We agree that a sensitivity study to ice physics, in particular basal
    sliding, is one of many improvements needed upon our results. However, for
    the kind of exercise presented here, the sensitivity to palaeo-climate
    history, which is virtually unknown and has many degrees of freedom, far
    outweight the model sensitivity to ice physics. From the perspective of
    glacial geology, a 5\,K change in air temperature is more influential than
    an order-of-magnitude increase in ice velocities (perhaps with the
    exception of our conclusions on trimlines, see below). In its current
    state, our manuscript studies the model sensitivity to the primary source
    of uncertainty, which is climate.

    Because the simulations presented here depend on computing
    resources only available through a peer-review application process, we are
    currently unable to perform an additional sensitivity study on basal
    sliding for the entire Alps. However, a regional study on the model
    sentivity to basal sliding is actually
    the topic for a follow-up study currently under preparation for the Rhine
    Glacier \citep{Imhof.etal.2017}, which has a relatively isolated catchment
    for its size as compared to other Alpine Glaciers.

    Meanwhile, the following text was added in the section on trimlines, which
    is certainly the most sensitive aspect of our results to basal sliding:

@@

    \msquote{%
        The remaining discrepancies may relate to temporal migrations of the
        basal thermal boundary, an absence of sliding in warm-based areas, and
        levelling of small-scale topographic features in the 1\,km horizontal
        grid. They call for more detailed comparisons spanning the entire
        Alpine range and specific sensitivity studies to relevant basal sliding
        and ice rheological parameters, and to the uncertain subglacial
        topography.}

    In addition, basal sliding and other sources of uncertainties in ice
    physics where also mentioned in the conclusions:

    \msquote{%
        However, these results are limited by uncertainties on ice physics. The
        till deformation model used here does not hold for sliding over bedrock
        surfaces. On the other hand, the constant friction angle used is
        representative of wet till but weaker basal conditions may have applied
        over saturated lake sediments where they occured. In the absence of ice
        deformation measuremets, a constant rheology was used for temperature
        ice containing more than 1\,\% of liquid water.}

@@

    We have now defined this better in the introduction:

    \msquote{
        During the last 800\,ka (thousand years before present)}

    \referee{%
        Could the reason for too large ice volumes when using the GRIP record,
        as mentioned in lines 13-15 on page 9 be due to Arctic Amplification?
        Could that have an effect then as it has now? This is also mentioned in
        line 11 on page 10.}

    These passages refer to the Younger Dryas and Dansgaard-Oeschger events.
    One must keep in mind that a linear scaling was applied to the GRIP record,
    so that it is already ``de-amplified'' before being used as temperature
    forcing. But still, Arctic amplification could be an explanation if it was stronger
    then than for changes that lead the LGM climate. Although we lack expertise
    on this topic, we could not find literature supporting this idea. Instead,
    current theories for the Younger Dryas and Dansgaard-Oeschger events mostly
    involve large amounts of freshwater discharge into the North Atlantic
    implying a more regional mechanism than Arctic amplification. To avoid
    restrictive interpretations, we thus decided not to mention Arctic
    amplification as a possible explanation.

@@

    \referee{\textbf{p.~6, l.~27:}
        replace ``is'' with ``are''.}

    Done.

    \referee{\textbf{p.~6, footnote:}
        add a reference for the correct formula for the rigidity and clarify
        (add something like, the consequence of this error is that the
        simulations effectively use....) Delete ``in'' before ``a small'' and
        add a quantification of the small change in the length scale – how
        small, is it a few percentage?}

    The correct formula can be found in the study by \citet[p.~443]{Love.1906}.
    The lenght scale of bedrock deformation can be computed as
    ${\alpha=\sqrt[4]{\frac{4D}{(\rho_m-\rho_l)g}}}$ \citep{Walcott.1970}, thus
    the error in lenght scale can be expressed as $\sqrt[4]{1+\nu}-1 = 5.7\%$.
    The footnote was clarified and references were added.

@@

    \referee{\textbf{p.~22, l.~6:}
        add ``records'' after forcing.}

    We replaced ``forcing'' by ``records''.

    \referee{\textbf{p.~22, l.~19:}
        why do you add ``potentially'' here? Isn't this a firm conclusion from
        you study?}

    The timing of the LGM extent depends on the actual climate history of the
    European Alps, which was almost certainly different and more complex than
    our climate forcing scenarii.  Although our simulations show that even a
    spatially homogeneous forcing produces spatial variations in the timing of
    maximum extent, on can not exclude that a different climate history might
    counteract these variations.

@@

    \referee{\textbf{p.~22, l.~20:}
        suggest to replace ``higher'' with ``larger''.}

    Done.

    \referee{\textbf{p.~22, l.~22:}
        same as above, why ``potentially'' here?}

    We removed the first ``potentially''. Nevertheless, the actual number and
    timing of glacier advances onto the foreland depends on Alpine climate
    history, which may have been very different from the forcing used in
    our simulations.

@@

    \referee{\textbf{p.~23, l.~2:}
        replace ``mode'' with ``more''.}

    Done.

    \referee{%
        Figures are generally clear and well set up.  The surface contours in
        Figs 4,5 and 6 is not clearly visible in my printout and could maybe
        become clearer?}

    Thank you. The 200\,m contour levels were thickened on all figures. We
    thank you again very much for the time and effort you put into this very
    meticulous review.

% ----------------------------------------------------------------------
% References

### 2018-07-23 articles/cycle/alps-cycle.tex
Reread the responses to referees 1, 2, and 3.

    Here, we use the Parallel Ice Sheet Model (PISM) to model the entire last
    glacial cycle (120--0\,ka) in the Alps, using horizontal resolutions of 2
    and 1\,km. Climate forcing is derived using present-day climate data from
    WorldClim and the ERA-Interim reanalysis, and time-dependent temperature
    offsets from multiple palaeo-climate proxies, among which only the EPICA ice
    core record yields glaciation during marine oxygen isotope stages~4
    (69--62\,ka) and~2 (34--18\,ka) spatially and temporally consistent with
    the geological reconstructions, while the other records used result in
    excessive early glacial cycle ice cover and a late Last Glacial Maximum.
    Despite the low variability of this Antarctic-based climate forcing, our
    simulation depicts a highly dynamic ice sheet, showing that Alpine glaciers
    may have advanced many times over the foreland during the last glacial
    cycle. Ice flow patterns during peak glaciation are largely governed by
    subglacial topography but include occasional transfluences and
    self-sustained ice domes. Modelled maximum ice surface is 861\,m
    higher than observed trimline elevations in the upper Rhone Valley, yet
    our simulation predicts little erosion at high elevation due to cold-based
    ice. Finally, despite the uniform climate forcing, differences in glacier
    catchment hypsometry cause the Last Glacial Maximum advance to be modelled
    as a time-transgressive event, with some glaciers reaching their maximum as
    early as 26\,ka, and others as late as 20\,ka.

@@

    Ice deformation is governed by the constitutive law for ice
    \citep{Glen.1952, Nye.1953}. Ice softness depends on ice temperature,
    pressure, and water content, through an enthalpy scheme
    \citep[Table~\ref{tab:params};][]{Aschwanden.etal.2012}. It follows a
    piece-wise Arrhenius-type rheology representative of Holocene polar ice
    \citep[Table~\ref{tab:params};][p.~77]{Cuffey.Paterson.2010}, and increases
    with liquid water fractions up to 0.01 \citep[p.~65--66]{Duval.1977,
    Lliboutry.Duval.1985, Cuffey.Paterson.2010}, an arbitrary threshold above
    which new measurements are critically needed \citep{Kleiner.etal.2015}.

@@

    Finally, in some simulations (hereafter labelled PP), precipitation was
    reduced with air temperature in
    order to simulate the potential rarefaction of atmospheric moisture in
    colder climates. This was done using an empirical relationship derived from
    observed accumulation rates and oxygen isotopes concentrations in the GRIP
    ice core \citep{Dahl-Jensen.etal.1993},
    %
    \begin{align}
      {\Psi}_{\text{PP}}(t) = \exp[\psi{\Delta}T_{\text{TS}}(t)] \,,
    \end{align}
    %
    with $\psi=0.169/2.4=0.0704$ \citep{Huybrechts.2002}. The control set of
    simulations use constant precipitation, corresponding to $\psi=0$.
    This simple relationship does not reflect the complexity of atmospheric
    circulation changes that governed moisture availability over the Alps
    during the last glacial cycle. Palaeoclimate proxies indicate slightly
    reduced LGM precipitation in western Europe with anomalies diminishing
    eastwards \citep{Wu.etal.2007}. Regional circulation models indicate
    generally dryer conditions during MIS~3 \citep{Barron.Pollard.2002,
    Kjellstrom.etal.2010} but more precipitation south of the Alps during MIS~2
    LGM \citep{Strandberg.etal.2011, Ludwig.etal.2016}. Besides, changes in the
    ice surface topography may have redistributed orographic precipitation.

@@

    In the upper Rhone Valley, observed trimlines are often located near the
    LGM cold-temperate basal thermal transition, or in cold-based areas
    (Fig.~\ref{fig:trimlines}c). The remaining discrepancies may relate to
    temporal migrations of the basal thermal boundary, an absence of sliding in
    warm-based areas, and levelling of small-scale topographic features in the
    1\,km horizontal grid. They call for more detailed comparisons
    spanning the entire Alpine range and specific sensitivity studies to
    relevant basal sliding and ice rheological parameters, and to the uncertain
    subglacial topography. Unfortunately, validation through the bedrock uplift
    rates \citep[cf.][]{Kuchar.etal.2012} is not doable in the Alps due its
    lower values, active tectonics and uncertainties on geological properties
    \citep[cf.][]{Mey.etal.2016}. However, the presence of an upper layer of
    cold ice during the LGM, already found at high altitude in the much warmer
    climate of today \citep[e.g.,][]{Suter.etal.2001, Bohleber.etal.2017}, is
    inevitable \citep[Fig.~3]{Haeberli.Schluchter.1987}.
    In this context, our results challenge the assumption that
    Alpine trimlines mark the maximum upper ice surface elevation of the LGM
    ice cover and call for a more accurate estimation of the thickness of the
    upper layer of cold ice in the Alps.

@@

    However, these results are limited by uncertainties on ice physics. The
    till deformation model used here does not hold for sliding over
    bedrock surfaces. On the other hand, the constant friction angle used is
    representative of wet till but weaker basal conditions may have applied
    over saturated lake sediments where they occured. A constant rheology was
    used for temperature ice containing more than 1\,\% of liquid water.

>>>>

    Here, we use the Parallel Ice Sheet Model (PISM) to model the entire last
    glacial cycle (120--0\,ka) in the Alps, using horizontal resolutions of 2
    and 1\,km. Climate forcing is derived using present-day climate data from
    WorldClim and the ERA-Interim reanalysis, and time-dependent temperature
    offsets from multiple palaeo-climate proxies, among which only the EPICA ice
    core record yields glaciation during marine oxygen isotope stages~4
    (69--62\,ka) and~2 (34--18\,ka) spatially and temporally consistent with
    the geological reconstructions, while the other records used result in
    excessive early glacial cycle ice cover and a late Last Glacial Maximum.
    Despite the low variability of this Antarctic-based climate forcing, our
    simulation depicts a highly dynamic ice sheet, showing that Alpine glaciers
    may have advanced many times over the foreland during the last glacial
    cycle. Ice flow patterns during peak glaciation are largely governed by
    subglacial topography but include occasional transfluences and
    self-sustained ice domes. Modelled maximum ice surface is 861\,m
    higher than observed trimline elevations in the upper Rhone Valley, yet
    our simulation predicts little erosion at high elevation due to cold-based
    ice. Finally, despite the uniform climate forcing, differences in glacier
    catchment hypsometry cause the Last Glacial Maximum advance to be modelled
    as a time-transgressive event, with some glaciers reaching their maximum as
    early as 27\,ka, and others as late as 21\,ka.

@@

    Ice deformation is governed by the constitutive law for ice
    \citep{Glen.1952, Nye.1953}. Ice softness depends on ice temperature,
    pressure, and water content, through an enthalpy scheme
    \citep[Table~\ref{tab:params};][]{Aschwanden.etal.2012}. It follows a
    piece-wise Arrhenius-type rheology representative of Holocene polar ice
    \citep[Table~\ref{tab:params};][p.~77]{Cuffey.Paterson.2010}, and increases
    with liquid water fractions up to 0.01 \citep[p.~65--66]{Duval.1977,
    Lliboutry.Duval.1985, Cuffey.Paterson.2010}, an arbitrary threshold above
    which new ice deformation measurements are critically needed
    \citep{Kleiner.etal.2015}.

@@

    Finally, in some simulations (hereafter labelled PP), precipitation was
    reduced with air temperature in
    order to simulate the potential rarefaction of atmospheric moisture in
    colder climates. This was done using an empirical relationship derived from
    observed accumulation rates and oxygen isotopes concentrations in the GRIP
    ice core \citep{Dahl-Jensen.etal.1993},
    %
    \begin{align}
      {\Psi}_{\text{PP}}(t) = \exp[\psi{\Delta}T_{\text{TS}}(t)] \,,
    \end{align}
    %
    with $\psi=0.169/2.4=0.0704$ \citep{Huybrechts.2002}. The control set of
    simulations use constant precipitation, corresponding to $\psi=0$.
    This simple relationship does not reflect the complexity of atmospheric
    circulation changes that governed moisture availability over the Alps
    during the last glacial cycle. Palaeoclimate proxies indicate slightly
    reduced LGM precipitation in western Europe with anomalies diminishing
    eastwards \citep{Wu.etal.2007}. Regional circulation models indicate
    generally dryer conditions during MIS~3 \citep{Barron.Pollard.2002,
    Kjellstrom.etal.2010} but more precipitation south of the Alps during MIS~2
    \citep{Strandberg.etal.2011, Ludwig.etal.2016}. Besides, changes in the
    ice surface topography may have redistributed orographic precipitation.

@@

    In the upper Rhone Valley, observed trimlines are often located near the
    LGM cold-temperate basal thermal transition, or in cold-based areas
    (Fig.~\ref{fig:trimlines}c). The remaining discrepancies may relate to
    temporal migrations of the basal thermal boundary, an absence of sliding in
    warm-based areas, and levelling of small-scale topographic features in the
    1\,km horizontal grid. They call for more detailed comparisons
    spanning the entire Alpine range and specific sensitivity studies to
    relevant basal sliding and ice rheological parameters, and to the uncertain
    subglacial topography. Unfortunately, validation through the bedrock uplift
    rate \citep[cf.][]{Kuchar.etal.2012} is not doable in the Alps due its
    lower values, active tectonics and uncertainties on geological properties
    \citep[cf.][]{Mey.etal.2016}. However, the presence of an upper layer of
    cold ice during the LGM, already found at high altitude in the much warmer
    climate of today \citep[e.g.,][]{Suter.etal.2001, Bohleber.etal.2017}, is
    inevitable \citep[Fig.~3]{Haeberli.Schluchter.1987}.
    In this context, our results challenge the assumption that
    Alpine trimlines mark the maximum upper ice surface elevation of the LGM
    ice cover and call for a more accurate estimation of the thickness of the
    upper layer of cold ice in the Alps.

@@

    However, these results are limited by uncertainties on ice physics. The
    till deformation model used here does not hold for sliding over
    bedrock surfaces. On the other hand, the constant friction angle used is
    representative of wet till but weaker basal conditions may have applied
    over saturated lake sediments where they occured. In the absence of ice
    deformation measurements, a constant rheology was used for temperature ice
    containing more than 1\,\% of liquid water.
